---
title: "Kentucky Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kentucky Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(kyschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6")
```

```{r fetch-data, cache = TRUE}
# Fetch data for visualizations
enr_2024 <- fetch_enr(2024, use_cache = TRUE)
enr <- fetch_enr_multi(2020:2024, use_cache = TRUE)
```

## 1. Jefferson County is Kentucky's giant

Jefferson County Public Schools (Louisville) serves 103,000 students, about 15% of Kentucky's entire enrollment. It's larger than the next five districts combined.

```{r story-1}
s1 <- enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(desc(n_students)) %>%
  select(district_name, n_students) %>%
  head(10)
stopifnot(nrow(s1) > 0)
s1
```

```{r jefferson-county}
top_districts <- enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, n_students))
stopifnot(nrow(top_districts) > 0)

print(top_districts %>% select(district_name, n_students))

ggplot(top_districts, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Kentucky's Largest School Districts",
       subtitle = "Jefferson County is nearly 15% of state enrollment",
       x = "", y = "Students") +
  theme_readme()
```

## 2. Kentucky enrollment has held steady near 686,000

Kentucky enrolled 698,000 students in 2020 and 686,000 in 2024 -- a modest 1.7% decline over five years. Despite COVID disruptions, enrollment has been remarkably stable.

```{r story-2}
s2 <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
stopifnot(nrow(s2) > 0)
s2
```

```{r enrollment-decline}
state_trend <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment")
stopifnot(nrow(state_trend) > 0)

print(state_trend %>% select(end_year, n_students))

ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020.7, y = max(state_trend$n_students), label = "COVID",
           color = "red", alpha = 0.7, hjust = 0, size = 3.5) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Kentucky Public School Enrollment",
       subtitle = "Stable near 686,000 students since 2020",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 3. Eastern Kentucky coal counties are shrinking

Appalachian coal counties have seen significant enrollment declines. Pike, Floyd, Letcher, and Perry counties combined had over 22,000 students in 2020, dropping to about 20,200 by 2024.

```{r story-3}
appalachian <- c("Pike County", "Floyd County", "Letcher County", "Perry County")

s3 <- fetch_enr_multi(2020:2024, use_cache = TRUE) %>%
  filter(grepl(paste(appalachian, collapse = "|"), district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, district_name, n_students)
stopifnot(nrow(s3) > 0)
s3
```

```{r appalachia-decline}
appalachia <- enr %>%
  filter(is_district, grepl(paste(appalachian, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")
stopifnot(nrow(appalachia) > 0)

print(appalachia)

ggplot(appalachia, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Eastern Kentucky Coal Counties",
       subtitle = "Pike, Floyd, Letcher, and Perry counties combined",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 4. Hispanic students now 10% of enrollment

Hispanic students grew from 7.7% to over 10% of enrollment between 2020 and 2024. Lexington, Louisville, and central Kentucky drive this growth.

```{r story-4}
s4 <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "hispanic") %>%
  select(end_year, n_students, pct)
stopifnot(nrow(s4) > 0)
s4
```

```{r hispanic-growth}
hispanic <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "hispanic")
stopifnot(nrow(hispanic) > 0)

print(hispanic %>% select(end_year, n_students, pct))

ggplot(hispanic, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["hispanic"]) +
  geom_point(size = 3, color = colors["hispanic"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  labs(title = "Hispanic Student Population in Kentucky",
       subtitle = "From 7.7% to over 10% since 2020",
       x = "School Year", y = "Percent of Students") +
  theme_readme()
```

## 5. COVID dipped enrollment by 15,000 students

Kentucky lost over 15,000 students between 2020 and 2021. The state recovered modestly in 2022-2023 before a slight decline in 2024.

```{r story-5}
s5 <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment",
         end_year %in% 2020:2024) %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students))
stopifnot(nrow(s5) > 0)
s5
```

```{r covid-impact}
covid_years <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment")
stopifnot(nrow(covid_years) > 0)

print(covid_years %>% select(end_year, n_students))

ggplot(covid_years, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020.7, y = max(covid_years$n_students), label = "COVID",
           color = "red", alpha = 0.7, hjust = 0, size = 3.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID's Impact on Kentucky Enrollment",
       subtitle = "Lost over 15,000 students in 2021, then gradual recovery",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 6. Fayette County holds steady alongside Louisville

While Jefferson County has remained relatively stable around 103,000, Fayette County (Lexington) has held at about 44,000 students. Both districts show resilience despite statewide pressures.

```{r story-6}
s6 <- enr %>%
  filter(grepl("Fayette County|Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, district_name, n_students) %>%
  tidyr::pivot_wider(names_from = district_name, values_from = n_students)
stopifnot(nrow(s6) > 0)
s6
```

```{r fayette-jefferson}
urban <- enr %>%
  filter(grepl("Fayette County|Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment")
stopifnot(nrow(urban) > 0)

print(urban %>% select(end_year, district_name, n_students))

ggplot(urban, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Kentucky's Two Urban Giants",
       subtitle = "Both districts relatively stable since 2020",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 7. 62% of students are economically disadvantaged

Kentucky has one of the highest rates of economic disadvantage in the nation. In some eastern Kentucky districts, over 90% of students qualify.

```{r story-7a}
s7a <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "econ_disadv") %>%
  select(n_students, pct)
stopifnot(nrow(s7a) > 0)
s7a
```

```{r story-7b}
# Highest rates
s7b <- enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "econ_disadv") %>%
  arrange(desc(pct)) %>%
  select(district_name, n_students, pct) %>%
  head(10)
stopifnot(nrow(s7b) > 0)
s7b
```

```{r econ-disadvantage}
econ_top <- enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "econ_disadv") %>%
  arrange(desc(pct)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, pct))
stopifnot(nrow(econ_top) > 0)

print(econ_top %>% select(district_name, pct))

ggplot(econ_top, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["black"]) +
  coord_flip() +
  labs(title = "Highest Rates of Economic Disadvantage",
       subtitle = "Eastern Kentucky districts often exceed 90%",
       x = "", y = "Percent Economically Disadvantaged") +
  theme_readme()
```

## 8. Kentucky is 71% white

Kentucky remains one of the less diverse states. Louisville and Lexington have significant minority populations; most rural districts are 90%+ white.

```{r story-8}
s8 <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  select(subgroup, n_students, pct) %>%
  arrange(desc(pct))
stopifnot(nrow(s8) > 0)
s8
```

```{r demographics}
demo <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  arrange(desc(pct)) %>%
  mutate(subgroup_label = reorder(subgroup, pct))
stopifnot(nrow(demo) > 0)

print(demo %>% select(subgroup, n_students, pct))

ggplot(demo, aes(x = subgroup_label, y = pct * 100)) +
  geom_col(aes(fill = subgroup)) +
  coord_flip() +
  scale_fill_manual(values = c("white" = colors["white"], "black" = colors["black"],
                               "hispanic" = colors["hispanic"], "asian" = colors["asian"],
                               "multiracial" = "#1ABC9C")) +
  labs(title = "Kentucky Student Demographics",
       subtitle = "State is 71% white",
       x = "", y = "Percent of Students") +
  theme_readme() +
  theme(legend.position = "none")
```

## 9. Boone County is Northern Kentucky's growth story

Boone County in the Cincinnati suburbs has about 21,500 students. It's now the third-largest district in Kentucky.

```{r story-9}
s9 <- enr %>%
  filter(grepl("Boone County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
stopifnot(nrow(s9) > 0)
s9
```

```{r boone-county}
boone <- enr %>%
  filter(is_district, grepl("Boone County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(boone) > 0)

print(boone %>% select(end_year, n_students))

ggplot(boone, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boone County - Northern Kentucky Growth",
       subtitle = "Cincinnati suburbs with stable enrollment",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 10. Independent districts are a Kentucky tradition

Kentucky has both county-wide districts (like Jefferson County) and independent city districts (like Bowling Green Independent). Some independent districts serve just a few hundred students.

```{r story-10}
s10 <- enr_2024 %>%
  filter(grepl("Independent", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(n_students) %>%
  select(district_name, n_students) %>%
  head(10)
stopifnot(nrow(s10) > 0)
s10
```

```{r independent-districts}
independent <- enr_2024 %>%
  filter(grepl("Independent", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(n_students) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, n_students))
stopifnot(nrow(independent) > 0)

print(independent %>% select(district_name, n_students))

ggplot(independent, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Kentucky's Smallest Independent Districts",
       subtitle = "Some serve just a few hundred students with their own school boards",
       x = "", y = "Students") +
  theme_readme()
```

## 11. Louisville is Kentucky's diversity hub

Jefferson County has significant Black (36%), Hispanic (18%), and Asian (5%) populations. White students make up only 35% of enrollment in Louisville.

```{r story-11}
s11 <- enr_2024 %>%
  filter(grepl("Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  select(subgroup, n_students, pct)
stopifnot(nrow(s11) > 0)
s11
```

```{r louisville-diversity}
# Louisville vs rest of state demographic comparison
jefferson <- enr_2024 %>%
  filter(grepl("Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  mutate(area = "Jefferson County (Louisville)")

state <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  mutate(area = "Kentucky Statewide")

compare <- bind_rows(jefferson, state) %>%
  mutate(subgroup = factor(subgroup, levels = c("white", "black", "hispanic", "asian", "multiracial")))
stopifnot(nrow(compare) > 0)

print(compare %>% select(area, subgroup, pct))

ggplot(compare, aes(x = subgroup, y = pct * 100, fill = area)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Jefferson County (Louisville)" = colors["black"],
                               "Kentucky Statewide" = colors["white"])) +
  labs(title = "Louisville vs Kentucky Demographics",
       subtitle = "Jefferson County is Kentucky's diversity hub",
       x = "", y = "Percent of Students", fill = "") +
  theme_readme()
```

## 12. The gender gap is minimal

Kentucky schools have nearly equal male and female enrollment, with males at 51.7% and females at 48.3%.

```{r story-12}
s12 <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) %>%
  select(subgroup, n_students, pct)
stopifnot(nrow(s12) > 0)
s12
```

```{r gender-balance}
gender <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) %>%
  mutate(subgroup = factor(subgroup, levels = c("male", "female")))
stopifnot(nrow(gender) > 0)

print(gender %>% select(subgroup, n_students, pct))

ggplot(gender, aes(x = subgroup, y = n_students, fill = subgroup)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")), vjust = -0.5, size = 5) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("male" = colors["white"], "female" = colors["hispanic"])) +
  labs(title = "Gender Balance in Kentucky Schools",
       subtitle = "Near-equal enrollment with slight male majority",
       x = "", y = "Students") +
  theme_readme() +
  theme(legend.position = "none")
```

## 13. Oldham County is the wealthy suburb

Oldham County, between Louisville and Lexington, has about 12,500 students. It has Kentucky's lowest economic disadvantage rate.

```{r story-13}
s13 <- enr %>%
  filter(grepl("Oldham County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
stopifnot(nrow(s13) > 0)
s13
```

```{r oldham-county}
oldham <- enr %>%
  filter(is_district, grepl("Oldham County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(oldham) > 0)

print(oldham %>% select(end_year, n_students))

ggplot(oldham, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["asian"]) +
  geom_point(size = 3, color = colors["asian"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Oldham County - The Wealthy Suburb",
       subtitle = "Stable enrollment between Louisville and Lexington",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 14. Harlan County tells the coal story

Harlan County dropped from 4,096 students in 2020 to 3,674 in 2024, a 10% decline in just five years. The county symbolizes the contraction of coal country in eastern Kentucky.

```{r story-14}
s14 <- fetch_enr_multi(2020:2024, use_cache = TRUE) %>%
  filter(grepl("Harlan County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
stopifnot(nrow(s14) > 0)
s14
```

```{r harlan-county}
harlan <- enr %>%
  filter(is_district, grepl("Harlan County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(harlan) > 0)

print(harlan %>% select(end_year, n_students))

ggplot(harlan, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["black"]) +
  geom_point(size = 3, color = colors["black"]) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Harlan County - The Coal Story",
       subtitle = "10% enrollment decline in five years",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 15. The multiracial population is growing

Multiracial students now make up 5.3% of Kentucky's enrollment, up from 4.4% in 2020.

```{r story-15}
s15 <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "multiracial") %>%
  select(end_year, n_students, pct)
stopifnot(nrow(s15) > 0)
s15
```

```{r multiracial-growth}
multiracial <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "multiracial")
stopifnot(nrow(multiracial) > 0)

print(multiracial %>% select(end_year, n_students, pct))

ggplot(multiracial, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = "#1ABC9C") +
  geom_point(size = 3, color = "#1ABC9C") +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Multiracial Students in Kentucky",
       subtitle = "5.3% of enrollment, up from 4.4% in 2020",
       x = "School Year", y = "Students") +
  theme_readme()
```

```{r session-info}
sessionInfo()
```
