---
title: "Kentucky Enrollment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kentucky Enrollment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(kyschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_readme <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6")
```

```{r fetch-data, cache = TRUE}
# Fetch data for visualizations
enr_2024 <- fetch_enr(2024, use_cache = TRUE)
enr <- fetch_enr_multi(2020:2024, use_cache = TRUE)
```

## 1. Jefferson County is Kentucky's giant

Jefferson County Public Schools (Louisville) serves 103,000 students, about 15% of Kentucky's entire enrollment. It's larger than the next five districts combined.

```{r story-1}
enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(desc(n_students)) %>%
  select(district_name, n_students) %>%
  head(10)
```

```{r jefferson-county}
top_districts <- enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, n_students))

ggplot(top_districts, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Kentucky's Largest School Districts",
       subtitle = "Jefferson County is nearly 15% of state enrollment",
       x = "", y = "Students") +
  theme_readme()
```

## 2. Kentucky enrollment peaked in 2020

Kentucky had nearly 1.5 million student enrollment records in 2020. The decline since has been dramatic.

```{r story-2}
enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
```

```{r enrollment-decline}
state_trend <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment")

ggplot(state_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Kentucky Public School Enrollment",
       subtitle = "Peaked in 2020, declining since",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 3. Eastern Kentucky coal counties are shrinking

Appalachian coal counties have seen significant enrollment declines. Pike, Floyd, Letcher, and Perry counties combined had over 22,000 students in 2020.

```{r story-3}
appalachian <- c("Pike County", "Floyd County", "Letcher County", "Perry County")

fetch_enr_multi(2020:2024, use_cache = TRUE) %>%
  filter(grepl(paste(appalachian, collapse = "|"), district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, district_name, n_students)
```

```{r appalachia-decline}
appalachia <- enr %>%
  filter(is_district, grepl(paste(appalachian, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

ggplot(appalachia, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Eastern Kentucky Coal Counties",
       subtitle = "Pike, Floyd, Letcher, and Perry counties combined",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 4. Hispanic students now 10% of enrollment

Hispanic students grew from 7.5% to over 10% of enrollment between 2020 and 2024. Lexington, Louisville, and central Kentucky drive this growth.

```{r story-4}
enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "hispanic") %>%
  select(end_year, n_students, pct)
```

```{r hispanic-growth}
hispanic <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "hispanic")

ggplot(hispanic, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["hispanic"]) +
  geom_point(size = 3, color = colors["hispanic"]) +
  labs(title = "Hispanic Student Population in Kentucky",
       subtitle = "From 7.5% to over 10% since 2020",
       x = "School Year", y = "Percent of Students") +
  theme_readme()
```

## 5. COVID hit Kentucky hard

Kentucky lost over 37,000 students between 2020 and 2021. The state saw some recovery in 2022-2023, but 2024 brought another significant decline.

```{r story-5}
enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment",
         end_year %in% 2020:2024) %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students))
```

```{r covid-impact}
covid_years <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "total_enrollment")

ggplot(covid_years, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID's Impact on Kentucky Enrollment",
       subtitle = "Lost over 37,000 students in 2021, partial recovery, then another decline",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 6. Fayette County holds steady while Louisville shrinks

While Jefferson County has remained relatively stable around 103,000, Fayette County (Lexington) has held at about 44,000 students.

```{r story-6}
enr %>%
  filter(grepl("Fayette County|Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, district_name, n_students) %>%
  tidyr::pivot_wider(names_from = district_name, values_from = n_students)
```

```{r fayette-jefferson}
urban <- enr %>%
  filter(grepl("Fayette County|Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment")

ggplot(urban, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Kentucky's Two Urban Giants",
       subtitle = "Both districts relatively stable since 2020",
       x = "School Year", y = "Students", color = "") +
  theme_readme()
```

## 7. 62% of students are economically disadvantaged

Kentucky has one of the highest rates of economic disadvantage in the nation. In some eastern Kentucky districts, over 90% of students qualify.

```{r story-7a}
enr_2024 %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "econ_disadv") %>%
  select(n_students, pct)
```

```{r story-7b}
# Highest rates
enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "econ_disadv") %>%
  arrange(desc(pct)) %>%
  select(district_name, n_students, pct) %>%
  head(10)
```

```{r econ-disadvantage}
econ_top <- enr_2024 %>%
  filter(is_district, grade_level == "TOTAL", subgroup == "econ_disadv") %>%
  arrange(desc(pct)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, pct))

ggplot(econ_top, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["black"]) +
  coord_flip() +
  labs(title = "Highest Rates of Economic Disadvantage",
       subtitle = "Eastern Kentucky districts often exceed 90%",
       x = "", y = "Percent Economically Disadvantaged") +
  theme_readme()
```

## 8. Kentucky is 71% white

Kentucky remains one of the less diverse states. Louisville and Lexington have significant minority populations; most rural districts are 90%+ white.

```{r story-8}
enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  select(subgroup, n_students, pct) %>%
  arrange(desc(pct))
```

```{r demographics}
demo <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  arrange(desc(pct)) %>%
  mutate(subgroup_label = reorder(subgroup, pct))

ggplot(demo, aes(x = subgroup_label, y = pct * 100)) +
  geom_col(aes(fill = subgroup)) +
  coord_flip() +
  scale_fill_manual(values = c("white" = colors["white"], "black" = colors["black"],
                               "hispanic" = colors["hispanic"], "asian" = colors["asian"],
                               "multiracial" = "#1ABC9C")) +
  labs(title = "Kentucky Student Demographics",
       subtitle = "State is 71% white",
       x = "", y = "Percent of Students") +
  theme_readme() +
  theme(legend.position = "none")
```

## 9. Boone County is Northern Kentucky's growth story

Boone County in the Cincinnati suburbs has about 21,500 students. It's now the third-largest district in Kentucky.

```{r story-9}
enr %>%
  filter(grepl("Boone County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
```

```{r boone-county}
boone <- enr %>%
  filter(is_district, grepl("Boone County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(boone, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boone County - Northern Kentucky Growth",
       subtitle = "Cincinnati suburbs with stable enrollment",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 10. Independent districts are a Kentucky tradition

Kentucky has both county-wide districts (like Jefferson County) and independent city districts (like Bowling Green Independent). Some independent districts serve just a few students.

```{r story-10}
enr_2024 %>%
  filter(grepl("Independent", district_name, ignore.case = TRUE),
         grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(n_students) %>%
  select(district_name, n_students) %>%
  head(10)
```

```{r independent-districts}
independent <- enr_2024 %>%
  filter(grepl("Independent", district_name, ignore.case = TRUE),
         grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  arrange(n_students) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, n_students))

ggplot(independent, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Kentucky's Smallest Independent Districts",
       subtitle = "Some serve just a few students with their own school boards",
       x = "", y = "Students") +
  theme_readme()
```

## 11. Louisville is Kentucky's diversity hub

Jefferson County has significant Black (36%), Hispanic (18%), and Asian (5%) populations. White students make up only 35% of enrollment in Louisville.

```{r story-11}
enr_2024 %>%
  filter(grepl("Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian")) %>%
  select(subgroup, n_students, pct)
```

```{r louisville-diversity}
# Louisville vs rest of state demographic comparison
jefferson <- enr_2024 %>%
  filter(grepl("Jefferson County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  mutate(area = "Jefferson County (Louisville)")

state <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) %>%
  mutate(area = "Kentucky Statewide")

compare <- bind_rows(jefferson, state) %>%
  mutate(subgroup = factor(subgroup, levels = c("white", "black", "hispanic", "asian", "multiracial")))

ggplot(compare, aes(x = subgroup, y = pct * 100, fill = area)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Jefferson County (Louisville)" = colors["black"],
                               "Kentucky Statewide" = colors["white"])) +
  labs(title = "Louisville vs Kentucky Demographics",
       subtitle = "Jefferson County is Kentucky's diversity hub",
       x = "", y = "Percent of Students", fill = "") +
  theme_readme()
```

## 12. The gender gap is minimal

Kentucky schools have nearly equal male and female enrollment, with males at 51.7% and females at 48.3%.

```{r story-12}
enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) %>%
  select(subgroup, n_students, pct)
```

```{r gender-balance}
gender <- enr_2024 %>%
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) %>%
  mutate(subgroup = factor(subgroup, levels = c("male", "female")))

ggplot(gender, aes(x = subgroup, y = n_students, fill = subgroup)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct * 100, 1), "%")), vjust = -0.5, size = 5) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("male" = colors["white"], "female" = colors["hispanic"])) +
  labs(title = "Gender Balance in Kentucky Schools",
       subtitle = "Near-equal enrollment with slight male majority",
       x = "", y = "Students") +
  theme_readme() +
  theme(legend.position = "none")
```

## 13. Oldham County is the wealthy suburb

Oldham County, between Louisville and Lexington, has about 12,500 students. It has Kentucky's lowest economic disadvantage rate.

```{r story-13}
enr %>%
  filter(grepl("Oldham County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
```

```{r oldham-county}
oldham <- enr %>%
  filter(is_district, grepl("Oldham County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(oldham, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["asian"]) +
  geom_point(size = 3, color = colors["asian"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Oldham County - The Wealthy Suburb",
       subtitle = "Stable enrollment between Louisville and Lexington",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 14. Harlan County tells the coal story

Harlan County had about 4,100 students in 2020. The county symbolizes the decline of coal country in eastern Kentucky.

```{r story-14}
fetch_enr_multi(2020:2024, use_cache = TRUE) %>%
  filter(grepl("Harlan County", district_name, ignore.case = TRUE),
         is_district, grade_level == "TOTAL", subgroup == "total_enrollment") %>%
  select(end_year, n_students)
```

```{r harlan-county}
harlan <- enr %>%
  filter(is_district, grepl("Harlan County", district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

ggplot(harlan, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["black"]) +
  geom_point(size = 3, color = colors["black"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Harlan County - The Coal Story",
       subtitle = "Enrollment in decline with coal country",
       x = "School Year", y = "Students") +
  theme_readme()
```

## 15. The multiracial population is growing

Multiracial students now make up 5.4% of Kentucky's enrollment, up from 4.3% in 2020.

```{r story-15}
enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "multiracial") %>%
  select(end_year, n_students, pct)
```

```{r multiracial-growth}
multiracial <- enr %>%
  filter(is_state, grade_level == "TOTAL", subgroup == "multiracial")

ggplot(multiracial, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = "#1ABC9C") +
  geom_point(size = 3, color = "#1ABC9C") +
  scale_y_continuous(labels = comma) +
  labs(title = "Multiracial Students in Kentucky",
       subtitle = "5.4% of enrollment, up from 4.3% in 2020",
       x = "School Year", y = "Students") +
  theme_readme()
```

```{r session-info}
sessionInfo()
```
